Does stomatal patterning in amphistomatous leaves minimize the CO\(_2\) diffusion path length with leaves of \emph{Arabidopsis thaliana}?

### 2023-11-10 ms/ms.tex
knitted new draft

% Create labels for Addresses if the are given in Elsevier format
   \MyNewLabel{ADDRTXT}{UHM}{%
  School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI
96822%
 }
   \MyNewLabel{ADDRTXT}{CU}{%
  Ecology and Evolutionary Biology, Universtity of Colorado, Boulder, CO
80309%
 }
   \MyNewLabel{ADDRTXT}{NIAB}{%
  Department of Crop Science and Production Systems, NIAB, Cambridge,
CB3 0LE, UK%
 }
   \MyNewLabel{ADDRTXT}{UCD}{%
  Department of Plant Sciences, University of California, Davis, CA
95616%
 }
   \MyNewLabel{ADDRTXT}{UWM}{%
  Department of Botany, University of Wisconsin, Madison, WI 53706%
 }

@@

\abstract{
We will write abstract after Discussion and Conclusions are complete}

@@

Stomatal anatomy (e.g.~size, density, distribution, and patterning) and
movement regulate gas exchange during photosynthesis, namely CO\(_2\)
assimilation and water loss through transpiration. Since waxy cuticles
are mostly impermeable to CO\(_2\) and H\(_2\)O, stomata are the primary
entry points through which gas exchange occurs despite making up a small
percentage of the leaf area \citep{lange_responses_1971}. Stomata
consist of two guard cells which open and close upon changes in turgor
pressure or hormonal cues \citep{mcadam_linking_2016}. The stomatal pore
leads to an internal space known as the substomatal cavity where gases
contact the mesophyll. Once in the mesophyll, CO\(_2\) diffuses
throughout a network of intercellular air space (IAS) and into mesophyll
cells where CO\(_2\) assimilation (\(A\)) occurs within the chloroplasts
\citep{lee_diffusion_1964}. Stomatal conductance and transpiration are
determined by numerous environmental and anatomical parameters such as
vapor pressure deficit (VPD), irradiance, temperature, wind speed, leaf
water potential, IAS geometry, mesophyll cell anatomy, and stomatal
anatomy.

Many successful predictions about stomata and other leaf traits can be
made by hypothesizing that natural selection should optimizes CO\(_2\)
gain per unit of water loss
\citep{cowan_stomatal_1977, buckley_optimal_2017, sperry_predicting_2017}.
However, stomatal anatomy may be partially constrained by physical and
developments limits on phenotypic expression
\citep{croxdale_stomatal_2000, harrison_influence_2020, muir_how_2023}.
Sometimes optimization leads to similar phenotypes across many disparate
species. For example, almost all stomata follow the one cell spacing
rule to maintain proper stomatal functioning
\citep{geisler_oriented_2000, dow_physiological_2014}; however some
species (notably in \emph{Begonia}) appear to benefit from overlapping
vapor shells caused by stomatal clustering
\citep{yi_gan_stomatal_2010, lehmann_effects_2015, papanatsiou_stomatal_2017}.
Stomatal traits also vary adaptively in different environments. Stomatal
density positively co-varies with irradiance during leaf development and
negatively co-varies with CO\(_2\) concentration
\citep{gay_influence_1975, schoch_dependence_1980, woodward_stomatal_1987, royer_stomatal_2001},
consistent with optimality predictions. Stomatal size is jointly
controlled by genome size, light, and stomatal density
\citep{jordan_environmental_2015}. Size positively co-varies with genome
size \citep{roddy_scaling_2020} and negatively co-varies with stomatal
density \citep{camargo_density_2011}. Total stomatal area (size
\(\times\) density) is optimized for operational conductance
(\(g_\text{s,op}\)) rather than maximum conductance (\(g_\text{s,max}\))
such that stomatal apertures are most responsive to changes in the
environment at their operational aperture
\citep{franks_physiological_2012, liu_scaling_2021}. Stomatal aperture
can compensate for maladaptive stomatal densities to an extent
\citep{bussis_stomatal_2006}, but stomatal density and size ultimately
determine a leaf's theoretical \(g_\text{s,max}\)
\citep{sack_developmental_2016}, which is proportional to
\(g_\text{s,op}\) \citep{murray_consistent_2020}. Additionally, low
stomatal densities lead to irregular and insufficient CO\(_2\) supply
and reduced photosynthetic efficiency in areas far from stomata
\citep{pieruschka_lateral_2006, morison_lateral_2005}, while high
stomatal densities can reduce water use efficiency (WUE)
\citep{bussis_stomatal_2006} and incur excessive metabolic costs
\citep{deans_optimization_2020}. In most species, stomata occur on the
abaxial (usually lower) leaf surface; but amphistomy, the occurrence of
stomata on both abaxial and adaxial leaf surfaces, is also prevalent in
high light environments with constant or intermittent access to
sufficient water
\citep{mott_adaptive_1982, jordan_using_2014, muir_light_2018, drake_two_2019, muir_is_2019}.
Amphistomy effectively halves the CO\(_2\) diffusion path length and
boundary layer resistance by doubling boundary layer conductance
\citep{parkhurst_adaptive_1978, harrison_influence_2020, mott_amphistomy_1991}.
Historically, stomatal patterning in dicot angiosperms was thought to be
random with an exclusionary distance surrounding each stomate
\citep{sachs_developmental_1974}; however, the developmental controls of
stomatal patterning are poorly understood and likely more complex than
random development along the leaf surface.
\citet{croxdale_stomatal_2000}{]} reviews three developmental theories
which attempt to explain stomatal patterning in angiosperms: inhibition,
cell lineage, and cell cycle, ultimately arguing for a cell cycle based
control of stomatal patterning.

The patterning and spacing of stomata on the leaf affects photosynthesis
in \(C_3\) leaves by altering the CO\(_2\) diffusion path length from
stomata to sites of carboxylation in the mesophyll. Maximum
photosynthetic rate (\(A_\text{max}\)) in \(C_3\) plants is generally
co-limited by biochemistry and diffusion, but modulated by light
availability
\citep{parkhurst_intercellular_1990, manter_ci_2004, carriqui_diffusional_2015}.
Low light decreasing CO\(_2\) demand by limiting electron transport
rate, leading to relatively high internal CO\(_2\) concentration
(\(C_\text{i}\)) and low \(A_\text{max}\) \citep{kaiser_metabolic_2016}.
In contrast, well hydrated leaves with open stomata in high light,
photosynthesis is often limited by CO\(_2\) supply as resistances from
the boundary layer, stomatal pore, and mesophyll can result in
insufficient CO\(C_2\) supply at the chloroplast to maxmimize
photosynthesis \citep{farquhar_biochemical_1980, lehmeier_cell_2017}. In
this study, we focus primarily on how stomatal patterning affects
diffusion, ignoring boundary layer and mesophyll resistances.

@@

\noindent \(\mathrm{NNI}\) is calculated for each sample image as
described above; \(\text{median}(\mathrm{NNI_{random}})\) and
\(\text{median}(\mathrm{NNI_{uniform}})\) are calculated from the
synthetic data specific to each sample image as described above. We
tested whether light treatment affects \(\mathrm{DI}\) and stomatla
density (\(D_S\)) using analysis of variance (ANOVA).

@@

\hypertarget{stomatal-distribution-is-nonrandom-but-far-from-ideal}{%
\subsection{Stomatal distribution is nonrandom, but far from
ideal}\label{stomatal-distribution-is-nonrandom-but-far-from-ideal}}

@@

Stomatal spacing on \emph{A. thaliana} leaves partially supports our
overall hypothesis that natural selection minimizes the average distance
between stomata and chloroplasts, for a given overall stomatal density.
There are three nonmutually exclusive hypotheses for why several of our
predictions were wrong. First, our predictions are wrong becuase they
are based on overly simplistic assumptions about epidermal and mesophyll
anatomy. Second, natural selection may be constrained by developmental
processes that prevent phenotypes from reaching their adaptive optima.
Third, the benefit of some traits may be of too little consequence to
result in fitness differences large enough to be respond to selection.
We consider the plausibility of these alternative hypotheses below and
present ideas for future work to test them.

{[}this paragraph should discuss why our assumptions may be too
simplistic{]} We assume an idealized leaf epidermal and mesophyll
structure that is homogenous and unconstrained by other tradeoffs. Real
leaves not only provide pathways for CO\(_2\) diffusion, but must supply
water, intercept light, and deter herbivores and pathogens. These
competing interests results in nonuniform epidermal and mesophyll
structure that could alter our predictions about optimal stomatal
spacing. {[}explain some examples{]}

{[}this paragraph should discuss developmental constraint{]} No
developmental pathway exists to ensure the ideal placement of stomata on
the leaf. {[}add more explanation{]}

{[}this paragraph should discuss limits on natural selection{]} The gas
exchange model demonstrate little photosynthetic gain from
abaxial-adaxial stomatal coordination (\autoref{fig:model_summary}).
Even though lateral diffusion may limit photosynthesis
\citep{morison_lateral_2005}, the marginal gain from optimally
offsetting stomata is not sufficient to generate fitness differences
relative to the strength of genetic drift (i.e.~the drift-barrier). We
can similarly extrapolate that an ideal, equilateral triangular stomatal
spacing is only slightly better than a suboptimal overdispered pattern.
Explaining these observations as the result of weak selection is in
tension with the finding that stomatal size and zone positively covary,
which would suggest that small changes in lateral diffusion distance are
significant. As described above, the positive correlation between
stomatal size and zone may be explained by common developmental
processes rather than as an adaptation to maximize CO\(_2\) diffusion.

{[}discuss significance of our study and future directions{]} Our study
corroborates previous studies which demonstrate that stomata are
non-randomly distributed along the leaf surface as a result of
developmental mechanisms such as spatially biased arrest of stomatal
initials \citep{boetsch_arrest_1995}, oriented asymmetric cell division
\citep{geisler_oriented_2000}, and cell cycle controls
\citep{croxdale_stomatal_2000}. We do not investigate the potential
developmental pathways that influence stomatal dispersion in this study;
however, they are important to consider as these pathways could limit
plants from reaching the theoretical peak in the adaptive landscape:
uniform stomatal dispersion. Instead, as this study suggests, plants may
simply compensate with higher stomatal density and by fitting stomatal
size to the area that they supply with CO\(_2\). To understand why
stomata are not ideally dispersed, more modelling should be done to
estimate the fitness gain of stomatal dispersion. Additionally, genetic
manipulation studies should attempt to create mutants with clustered and
ideally dispersed stomata for a comparison of their photosynthetic
traits. This could have extremely important implications for maximum
assimilation rates in crops as most crop species are grown in high light
where CO\(_2\) is often limiting. In drought-prone environments,
increased stomatal dispersion may increase water use efficiency by
reducing the number of stomata needed to achieve the same internal
CO\(_2\) concentration, \(C_\text{i}\).

{[}Conclusion{]} Our results suggest that after optimiziing stomatal
density and having developmental rules for spacing stomata relatively
evenly, there may be limited gains to further optimization. Therefore,
developmental constraints may be necessary to make sense of some
features of stomatal spacing and distribution.

@@

JLW and CDM conceived of the project, analyzed data, and wrote the
manuscript. GW provided data. TNB contributed to model development and
helped edit the manuscript.

>>>>

% Create labels for Addresses if the are given in Elsevier format
   \MyNewLabel{ADDRTXT}{UHM}{%
  School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI
96822%
 }
   \MyNewLabel{ADDRTXT}{CU}{%
  Ecology and Evolutionary Biology, University of Colorado, Boulder, CO
80309%
 }
   \MyNewLabel{ADDRTXT}{NIAB}{%
  Department of Crop Science and Production Systems, NIAB, Cambridge,
CB3 0LE, UK%
 }
   \MyNewLabel{ADDRTXT}{UCD}{%
  Department of Plant Sciences, University of California, Davis, CA
95616%
 }
   \MyNewLabel{ADDRTXT}{UWM}{%
  Department of Botany, University of Wisconsin, Madison, WI 53706%
 }

@@

\abstract{
Photosynthesis is co-limited by multiple factors depending on the plant
and its environment. These include biochemical rate limitations,
internal and external water potentials, temperature, irradiance, and
carbon dioxide (CO2). Amphistomatous leaves have stomata on both abaxial
and adaxial leaf surfaces. This feature is considered an adaptation to
alleviate CO2 diffusion limitations in productive environments where
other factors are not limiting as the diffusion path length from stomate
to chloroplast is effectively halved. Plants can also reduce CO2
limitations through other aspects of optimal stomatal anatomy: stomatal
density, distribution and patterning, and size. A number of studies have
demonstrated that stomata are overdispersed on a single leaf surface;
however, much less is known about stomatal anatomy in amphistomatous
plants, especially the coordination between leaf surfaces, despite their
prevelance in nature and near ubiquity among crop species. Here we use
novel spatial statistics based on simulations and photosynthesis
modelling to test hypotheses about how amphistomatous plants may
optimize CO2 limitations in the model angiosperm Arabidopsis thaliana
grown in different light environments. We find that 1) stomata are
overdispersed, but not ideally dispersed, on both leaf surfaces across
all light treatments; 2) abaxial and adaxial leaf surfaces are
independent; and 3) the theoretical improvements to photosynthesis from
abaxial-adaxial stomatal coordination are miniscule (\textasciitilde1\%)
across the range of feasible traitspace. However, we also find that 4)
stomatal size is correlated with the mesophyll volume that it supplies
with CO2, suggesting that plants may optimize CO2 diffusion limitations
through alternative pathways other than ideal, uniform stomatal spacing.
We discuss the myriad developmental, physical, and evolutionary
constraits which may prohibit plants from reaching the theoretical
adaptive peak of uniform stomatal spacing and inter-surface stomatal
coordination. These findings contribute to our understanding of
amphistomatous plants' exploration of the leaf anatomy adaptive
landscape.}

@@

Stomatal anatomy (e.g.~size, density, distribution, and patterning) and
movement regulate gas exchange during photosynthesis, namely CO\(_2\)
assimilation and water loss through transpiration. Since waxy cuticles
are mostly impermeable to CO\(_2\) and H\(_2\)O, stomata are the primary
entry and exit points through which gas exchange occurs despite making
up a small percentage of the leaf area \citep{lange_responses_1971}.
Stomata consist of two guard cells which open and close upon changes in
turgor pressure or hormonal cues \citep{mcadam_linking_2016}. The
stomatal pore leads to an internal space known as the substomatal cavity
where gases contact the mesophyll. Once in the mesophyll, CO\(_2\)
diffuses throughout a network of intercellular air space (IAS) and into
mesophyll cells where CO\(_2\) assimilation (\(A\)) occurs within the
chloroplasts \citep{lee_diffusion_1964}. Stomatal conductance and
transpiration are determined by numerous environmental and anatomical
parameters such as vapor pressure deficit (VPD), irradiance,
temperature, wind speed, leaf water potential, IAS geometry, mesophyll
cell anatomy, and stomatal anatomy.

Many successful predictions about stomata and other C3 leaf traits can
be made by hypothesizing that natural selection should optimize CO\(_2\)
gain per unit of water loss for any given set of environmental
parameters, including their variability
\citep{cowan_stomatal_1977, buckley_optimal_2017, sperry_predicting_2017}.
Total stomatal area (size \(\times\) density) is optimized for
operational conductance (\(g_\text{s,op}\)) rather than maximum
conductance (\(g_\text{s,max}\)) such that stomatal apertures are most
responsive to changes in the environment at their operational aperture
\citep{franks_physiological_2012, liu_scaling_2021}. Stomatal aperture
can compensate for maladaptive stomatal densities to an extent
\citep{bussis_stomatal_2006}, but stomatal density and size ultimately
determine a leaf's theoretical \(g_\text{s,max}\)
\citep{sack_developmental_2016}, which is proportional to
\(g_\text{s,op}\) \citep{murray_consistent_2020}. Additionally, low
stomatal densities lead to irregular and insufficient CO\(_2\) supply
and reduced photosynthetic efficiency in leaf areas far from stomata
\citep{pieruschka_lateral_2006, morison_lateral_2005}, while high
stomatal densities can reduce water use efficiency (WUE)
\citep{bussis_stomatal_2006} and incur excessive metabolic costs
\citep{deans_optimization_2020}. Stomatal density positively co-varies
with irradiance during leaf development and negatively co-varies with
CO\(_2\) concentration
\citep{gay_influence_1975, schoch_dependence_1980, woodward_stomatal_1987, royer_stomatal_2001},
consistent with optimality predictions. In most species, stomata occur
only on the abaxial (usually lower) leaf surface; but amphistomy, the
occurrence of stomata on both abaxial and adaxial leaf surfaces, is also
prevalent in high light environments with constant or intermittent
access to sufficient water
\citep{mott_adaptive_1982, jordan_using_2014, muir_light_2018, drake_two_2019, muir_is_2019}.
Amphistomy effectively halves the CO\(_2\) diffusion path length and
boundary layer resistance by doubling boundary layer conductance
\citep{parkhurst_adaptive_1978, harrison_influence_2020, mott_amphistomy_1991}.
Despite these optimality predictions, stomatal anatomy may be partially
constrained by physical and developmental limits on phenotypic
expression
\citep{croxdale_stomatal_2000, harrison_influence_2020, muir_how_2023}.

A number of physical and developmental processes constrain stomatal
anatomy traitspace. For example, almost all stomata follow the one cell
spacing rule to maintain proper stomatal functioning as guard cell
movement requires the rapid exchange of ions with neighboring epidermal
cells (i.e.~subsidiary cells)
\citep{geisler_oriented_2000, dow_physiological_2014}. This would
prevent stomata from being strongly clustered; however, some species
(notably in \emph{Begonia}) appear to benefit from the overlapping vapor
shells caused by stomatal clustering in dry environments
\citep{yi_gan_stomatal_2010, lehmann_effects_2015, papanatsiou_stomatal_2017}.
Historically, stomatal patterning in dicot angiosperms was thought to be
random with an exclusionary distance surrounding each stomate
\citep{sachs_developmental_1974}; however, the developmental controls of
stomatal patterning are more complex than random development along the
leaf surface. \citet{croxdale_stomatal_2000} reviews three developmental
theories which attempt to explain stomatal patterning in angiosperms:
inhibition, cell lineage, and cell cycle, ultimately arguing for a cell
cycle based control of stomatal patterning.
\citet{pillitteri_mechanisms_2012} review the short and long distance
signalling pathways associated with stomatal spacing and development,
which include cell to cell communication and whole plant integration to
ensure the proper spacing of stomata across a single leaf surface
depending on environmental ques. Much less is known about the
development of stomata on the adaxial leaf surface in amphistomatous
plants. Stomatal size is additionally constrained by genome size with
larger genomes leading to larger minimum guard cell size
\citep{jordan_environmental_2015}. Despite these limitations,
ecophysiological theory still predicts optimal stomatal anatomy, the
details of which are discussed below.

The patterning and spacing of stomata on the leaf affects photosynthesis
in \(C_3\) leaves by altering the CO\(_2\) diffusion path length from
stomata to sites of carboxylation in the mesophyll. Maximum
photosynthetic rate (\(A_\text{max}\)) in \(C_3\) plants is generally
co-limited by biochemistry and diffusion, but modulated by light
availability
\citep{parkhurst_intercellular_1990, manter_ci_2004, carriqui_diffusional_2015}.
Low light decreases CO\(_2\) demand by limiting electron transport rate,
leading to relatively high internal CO\(_2\) concentration
(\(C_\text{i}\)) and low \(A_\text{max}\) \citep{kaiser_metabolic_2016}.
In contrast, well hydrated leaves with open stomata in high light,
photosynthesis is often limited by CO\(_2\) supply as resistances from
the boundary layer, stomatal pore, and mesophyll can result in
insufficient CO\(C_2\) supply at the chloroplast to maxmimize
photosynthesis \citep{farquhar_biochemical_1980, lehmeier_cell_2017}. In
this study, we focus primarily on how stomatal patterning affects
diffusion, ignoring boundary layer and mesophyll resistances.

@@

\noindent \(\mathrm{NNI}\) is calculated for each sample image as
described above; \(\text{median}(\mathrm{NNI_{random}})\) and
\(\text{median}(\mathrm{NNI_{uniform}})\) are calculated from the
synthetic data specific to each sample image as described above. We
tested whether light treatment affects \(\mathrm{DI}\) and stomatal
density (\(D_S\)) using analysis of variance (ANOVA).

@@

\hypertarget{stomatal-distribution-is-non-random-but-far-from-ideal}{%
\subsection{Stomatal distribution is non-random, but far from
ideal}\label{stomatal-distribution-is-non-random-but-far-from-ideal}}

@@

Stomatal spacing on \emph{A. thaliana} leaves partially supports our
overall hypothesis that natural selection minimizes the average distance
between stomata and chloroplasts, for a given overall stomatal density.
There are three non-mutually exclusive hypotheses for why several of our
predictions were wrong. First, our predictions are wrong becuase they
are based on overly simplistic assumptions about epidermal and mesophyll
anatomy. Second, natural selection may be constrained by developmental
processes that prevent phenotypes from reaching their adaptive optima.
Third, the benefit of some traits may be of too little consequence to
result in fitness differences large enough to respond to selection. We
consider the plausibility of these alternative hypotheses below and
present ideas for future work to test them.

We assume an idealized leaf epidermal and mesophyll structure that is
homogenous and unconstrained by other tradeoffs. Real leaves not only
provide pathways for CO\(_2\) diffusion, but must supply water,
intercept light, and deter herbivores and pathogens. These competing
interests result in nonuniform epidermal and mesophyll structure that
could alter our predictions about optimal stomatal spacing. In order to
maintain consistent leaf water potential across the lamina, stomatal
density must be coordinated with vein density
\citep{fiorin_transport_2016}. Thus, stomatal spacing may be optimized
not at the interstomatal level, but at a higher level, coordinating
water transport and water loss. We also assume a uniform mesophyll;
however, in ab-adaxially differentiated leaves, this assumption is
overly simplistic. The pallisade mesophyll is more tightly packed than
the spongy mesophyll as an adaptation to intercept light efficiently, so
lateral diffusion may be more limiting in the adaxial portion of the
leaf. This may explain why adaxial leaf surfaces have consistently
higher dispersion indices than abaxial surfaces across all light
treatments (\autoref{fig:single-surface}). Future gas exchange models
should include multiple parameter sets for different strata of mesophyll
tissue.

No developmental pathway exists to ensure an idealized placement of
stomata on the leaf surface. Rather, stomatal development is a dynamic
process that must be plastic to environmental ques. Leaves develop based
on short and long distance signalling pathways which relay information
about incoming light, humidity, temperature, and surrounding stomata to
developing leaf tissues \citep{pillitteri_mechanisms_2012}. Our results
showing an intermediate level of dispersion in stomatal spacing may be
best explained by these developmental pathways which ensure the proper
spacing of stomata, with an added random effect brought about by the
necessity for plasticity in stomatal development
(\autoref{fig:single-surface}). However, any nonideal randomness in
stomatal spacing may offset by simultaneous and coordinated development
of the IAS \citep{baillie_developmental_2020}. The fact that stomata
which supply a greater mesophyll volume tend to be larger suggests that
plants may use coordinated development of multiple leaf tissues to
negate nonideal stomatal spacing (\autoref{fig:length-area}).

In amphistomatous leaves, ideal stomatal spacing is complicated by a
third dimension. Our gas exchange model demonstrates little
photosynthetic gain from abaxial-adaxial stomatal coordination
(\autoref{fig:model_summary}). Even though lateral diffusion may limit
photosynthesis \citep{morison_lateral_2005}, the marginal gain from
optimally offsetting stomata is not sufficient to generate fitness
differences relative to the strength of genetic drift (i.e.~the
drift-barrier). We can similarly extrapolate that an ideal, equilateral
triangular stomatal spacing is only slightly better than a suboptimal
pattern. Any benefit garnered by ideal stomatal spacing may be
additionally offset by a cost to developmental flexibility in variable
environments
\citep{pillitteri_mechanisms_2012, baillie_developmental_2020}.
Explaining these observations as the result of weak selection is in
tension with the finding that stomatal size and zone positively covary,
which would suggest that small changes in lateral diffusion distance are
significant. As described above, the positive correlation between
stomatal size and zone may be explained by common developmental
processes rather than as an adaptation to maximize CO\(_2\) diffusion.
Such flexibility in mature stomate size is an alternate peak in the
adaptive landscape which cannot be assessed with our simplistic model
assumptions. In any case, there is no evidence for coordinated
development of both leaf surfaces, and very little theoretical benefit
to photosynthesis, except in marginal circumstances which are
exceptionally rare in nature.

Our study corroborates previous studies which demonstrate that stomata
are non-randomly distributed along the leaf surface as a result of
developmental mechanisms such as spatially biased arrest of stomatal
initials \citep{boetsch_arrest_1995}, oriented asymmetric cell division
\citep{geisler_oriented_2000}, and cell cycle controls
\citep{croxdale_stomatal_2000}. We do not investigate the potential
developmental pathways that influence stomatal dispersion in this study;
however, they are important to consider as these pathways could limit
plants from reaching a theoretical peak in the adaptive landscape:
uniform stomatal dispersion. Instead, as this study suggests, plants may
simply compensate with higher stomatal density and by fitting stomatal
size to the area that they supply with CO\(_2\). To understand why
stomata are not ideally dispersed, more modelling (with a more complex
set of assumptions including vein density and IAS structure) should be
done to estimate the fitness gain of stomatal dispersion. Additionally,
genetic manipulation studies should attempt to create mutants with
clustered and ideally dispersed stomata for a comparison of their
photosynthetic traits. This could have extremely important implications
for maximum assimilation rates in crops as most crop species are grown
in high light where CO\(_2\) is often limiting. In drought-prone
environments, increased stomatal dispersion may increase water use
efficiency by reducing the number of stomata needed to achieve the same
internal CO\(_2\) concentration, \(C_\text{i}\).

Our results suggest that after optimizing stomatal density and having
developmental rules for spacing stomata relatively evenly, there may be
limited gains to further optimization. Therefore, developmental
constraints may be necessary to make sense of some features of stomatal
spacing and distribution. The possibility that ideal stomatal spacing is
not the ``tallest'' fitness peak must also be explored, as stomate size
is demonstrated in this study to covary with mesophyll volume supplied
with CO\(_2\). This may be especially true in highly variable
environments or in large tree species with sun and shade leaves where
developmental ques may change rapidly. The temporal component, not
considered here, could also have significant implications, as CO\(_2\)
may only be limiting to photosynthesis during short periods when all
other conditions are ideal. In these cases, the theoretical benefits of
ideal stomatal spacing are further diminished. Future exploration of
these competing hypotheses would require more advanced modelling,
additional exploration of IAS space development and its effects on gas
exchange, both real and modelled, and knowledge about how often the
species of interest is CO\(_2\) limited across of range of natural
settings. Despite these additional considerations, this paper represents
an important contribution to understanding the potential drivers of and
limitations to stomatal anatomy in amphistomatous plants.

@@

JLW and CDM conceived of the project, analyzed data, and wrote the
manuscript. GJD provided data. TNB contributed to model development and
helped edit the manuscript.
